Fitting correlated interferometric data

### 2020-01-15 article.tex
details

\begin{equation} 
   \vcov = \begin{pmatrix} 
     \absdev^2\cot^2 + \reldev^2\data_1^2 & \reldev^2\data_1\data_2\\
     \reldev^2\data_1\data_2              & \absdev^2\cot^2 + \reldev^2\data_2^2
            \end{pmatrix},
\end{equation}
when the second-order error term ($\absdev\reldev$) is ignored.  Then, the least squares estimate for $\data$ 
\begin{equation}
    {\data} = \frac{\data_1 + \data_2}{2} 
            \left(1 + \reldev^2\frac{(\data_1-\data_2)^2}{2\tau^2\absdev^2}\right)^{-1}
\end{equation}
systematically falls below the average.  If measurements differ significantly, it can even fall below the lowest value.  Figure~\ref{fig:peelle} gives such an example with 
\begin{align*}
    \cot   &= 1.250 \pm 0.063\ (5\%),\\
    \raw_1 &= 0.634 \pm 0.004\ (0.6\%),\\
    \raw_2 &= 0.646 \pm 0.004\ (0.6\%),
\intertext{which yields two points 2.1 standard deviations appart}
    \data_1 &= \system{0.793}{0.005}{0.040},\\
    \data_2 &= \system{0.808}{0.005}{0.040}
\intertext{and the estimate}
    \data &= \system{0.791}{0.003}{0.040}
\end{align*}
falls outside the data range.
\begin{figure}
\includegraphics[width=\linewidth]{pdf/original-peelle.pdf}
\caption{Original Peelle problem rewritten in the context of interferometry.  Two correlated visibility measurements $\data_1$ and $\data_2$ (points with error bars) yield an estimate $\data$ (dashed line, uncertainty in grey) outside of the data range. Systematic uncertainty from calibration is 5\% and measurement noise $\approx 0.6\%$.}
\label{fig:peelle}
\end{figure}

@@

\begin{table}[t]
\caption{Symbols used in this paper. Lower case bold font is used for vectors and upper case bold font for matrices.}
\begin{tabular}{ll}
\hline\hline
Symbol               & Meaning\\
\hline
$\mean a$            & true value of $a$\\
$<a>$                & expectancy of $a$\\
$\tr{\vec A}$        & transpose of $\vec A$\\
$\vec c = \vec a\hadam\vec b$ & elementwise product of $\vec a$ and $\vec b$\\
$\vec C = \vec a\outer\vec b$ & outer product of $\vec a$ and $\vec b$\\
$\vec c = \vec A\vec b$       & matrix product of $\vec A$ and $\vec b$\\
$\delta$             & Kronecker delta\\
\hline
$\vdata$             & data\\ 
$\verror$            & error ($= \vdata - \vdatamean$)\\
$\sigma^2$           & variance ($= \expect{\error^2}$)\\
$\vcov$              & covariance matrix ($= \expect{\verror\outer\verror}$)\\
$\corr$              & correlation coefficient\\
$\vsens$, $\msens$   & sensitivity vector or matrix\\
$\vparam$            & parameters of the model\\
$\vmod$              & model values ($= \vsens\vparam\approx\vdatamean$)\\
\hline
$a\meas$             & $a$ of the measurement error\\
$a\norm$             & $a$ of the normalisation error\\
$\ppp a$             & $a$ impacted by PPP\\
\hline
\end{tabular}
\end{table}

>>>>

\begin{equation} 
   \vcov = \begin{pmatrix} 
     \absdev^2\cot^2 + \reldev^2\data_1^2 & \reldev^2\data_1\data_2\\
     \reldev^2\data_1\data_2              & \absdev^2\cot^2 + \reldev^2\data_2^2
            \end{pmatrix},
\end{equation}
when the second-order error term ($\absdev\reldev$) is ignored.  Using the matrix inverse $\vcov^{-1}$, the least squares estimate is 
\begin{equation}
\begin{split}
    {\data} &= \frac{  (\vcov^{-1}_{11} + \vcov^{-1}_{12}) \data_1
                      +(\vcov^{-1}_{22} + \vcov^{-1}_{12}) \data_2 }
                    {  \vcov^{-1}_{11} + \vcov^{-1}_{22} + 2\vcov^{-1}_{12}},\\
            &= \frac{\data_1 + \data_2}{2} 
            \left(1 + \reldev^2\frac{(\data_1-\data_2)^2}{2\tau^2\absdev^2}\right)^{-1}.
\end{split}
\end{equation}
It systematically falls below the average of the two values $\data_1$ and $\data_2$.  If measurements differ significantly, it can even fall below the lowest value.  Figure~\ref{fig:peelle} gives such an example with a instrumental visibility of 50\% and two measurements on an unresolved target: 
\begin{align*}
    \cot   &= 2.000 \pm 0.100\ (5\%),\\
    \raw_1 &= 0.495 \pm 0.003\ (\approx 0.6\%),\\
    \raw_2 &= 0.505 \pm 0.003\ (\approx 0.6\%),
\intertext{which yields two points 2.4 standard deviations appart}
    \data_1 &= \system{0.990}{0.006}{0.050},\\
    \data_2 &= \system{1.010}{0.006}{0.050}
\intertext{and the estimate}
    \data &= \system{0.986}{0.004}{0.049}
\end{align*}
falls outside the data range.
\begin{figure}[t]
\includegraphics[width=\linewidth]{pdf/original-peelle.pdf}
\caption{Original Peelle problem rewritten in the context of interferometry.  \emph{Bottom:} Two raw visibility amplitudes $\nu_1$ and $\nu_2$ (points with statistic error bars of $\approx 0.6\%$) are calibrated by the transfer function $1/\tau$ (solid line with systematic error zone of 5\%). \emph{Top:} the two calibrated visibility measurements $\data_1$ and $\data_2$ (points with statistic error bars of $\approx 0.6\%$) are strongly correlated. The least-squares estimate for the visibility $\data$ (dashed line, with statistic uncertainty error zone displayed) falls outside of the data range.}
\label{fig:peelle}
\end{figure}

@@

\begin{table}[t]
\caption{Symbols used in this paper. Lower case bold font is used for vectors and upper case bold font for matrices.}
\begin{tabular}{ll}
\hline\hline
Symbol               & Meaning\\
\hline
$\mean a$            & true value of $a$\\
$<a>$                & expectancy of $a$\\
$\tr{\vec A}$        & transpose of $\vec A$\\
$\vec c = \vec a\hadam\vec b$ & elementwise product of $\vec a$ and $\vec b$\\
$\vec C = \vec a\outer\vec b$ & outer product of $\vec a$ and $\vec b$\\
$\vec c = \vec A\vec b$       & matrix product of $\vec A$ and $\vec b$\\
$\delta$             & Kronecker delta\\
\hline
$\vdata$             & data\\ 
$\verror$            & error ($= \vdata - \vdatamean$)\\
$\dev^2$             & variance ($= \expect{\error^2}$)\\
$\vcov$              & covariance matrix ($= \expect{\verror\outer\verror}$)\\
$\corr$              & correlation coefficient\\
$\vsens$, $\msens$   & sensitivity vector or matrix\\
$\vparam$            & parameters of the model\\
$\vmod$              & model values ($= \vsens\vparam\approx\vdatamean$)\\
\hline
$a\meas$             & $a$ of the measurement error\\
$a\norm$             & $a$ of the normalisation error\\
$\ppp a$             & $a$ impacted by PPP\\
\hline
\end{tabular}
\end{table}
